---
title: "Monocle workflow"
author: "sharan"
date: "11/01/2021"
output: html_document
---

```{r setup, include=FALSE, results='hide'}
knitr::opts_chunk$set(echo = TRUE)
setwd(" ")

library(monocle3)
library(Seurat)
library(dplyr)

```

## Load files and make cds object
Expression matrix, annotation, and metadata were extracted from the Seurat object and stored as RDS to load.
Gene annotation needs to be a factor thus converting it.

```{r}
singlets_expression_matrix = readRDS(file = "singlet_expression_matrix.rds")
singlets_gene_annotation = readRDS(file = "singlet_gene_annotation.rds")
singlets_cell_metadata = readRDS(file = "singlet_cell_metadata.rds")
singlets_gene_annotation$gene_short_name = singlets_gene_annotation$gene_short_name %>% as.factor()


# Make the CDS object
cds <- new_cell_data_set(singlets_expression_matrix,
                         cell_metadata = singlets_cell_metadata,
                         gene_metadata = singlets_gene_annotation)


```

## Preprocess and check data

```{r}
cds <- preprocess_cds(cds, num_dim = 25)
plot_pc_variance_explained(cds)
```
25 dimensions seem more than enough, most variation is captured in first 10

```{r, message=FALSE, echo=FALSE}
cds <- reduce_dimension(cds)
plot_cells(cds, color_cells_by = "HTO_classification", 
           label_cell_groups = FALSE, reduction_method = "UMAP",
           cell_size = 1)
plot_cells(cds, genes = c("Aif1","Apoe"), cell_size = 1)
```
We reduce dimension for visualization here, the standard for monocle is UMAP, which is also better for trajectory analysis. we can then visualize the reduced data by the metadata, expression of genes (and even clusters and partitions after clustering is performed)

```{r, message=FALSE, echo=FALSE}
cds <- cluster_cells(cds, resolution = 0.000037 ) # above resolution = 0.000036 leads to 4 clusters but also 4 partitions, while below is 3 and 1 respectively

plot_cells(cds, color_cells_by = "cluster", 
           label_cell_groups = FALSE, cell_size = 1) 
plot_cells(cds, color_cells_by = "partition", 
           label_cell_groups = FALSE, cell_size = 1)
```



## Differential analysis

```{r, message=FALSE, echo=FALSE}

marker_test_res <- top_markers(cds, group_cells_by="cluster", 
                               reference_cells=1000)

top_specific_markers <- marker_test_res %>%
  filter(fraction_expressing >= 0.10) %>%
  group_by(cell_group) %>%
  top_n(3, pseudo_R2)

top_specific_marker_ids <- unique(top_specific_markers %>% pull(gene_id))

plot_genes_by_group(cds,
                    top_specific_marker_ids,
                    group_cells_by="cluster",
                    ordering_type="cluster_row_col",
                    max.size=8)

```



```{r, message=FALSE, echo=FALSE}

```


## Regression analysis

In this analysis, we are seeing if some interesting genes (such as Apoe Plcg2 etc) vary between clusters

```{r, message=FALSE, echo=FALSE}
interest_genes <- c("Apoe","Plcg2","Sod2","C1qa")
cds_subset <- cds[rowData(cds)$gene_short_name %in% interest_genes,]

gene_fits <- fit_models(cds_subset, model_formula_str = "~cluster")
fit_coefs <- coefficient_table(gene_fits)

cluster_terms <- fit_coefs %>% filter(term != "(Intercept)") #Removing intercept terms, we dont want them. (where is result for cluster1?)


cluster_terms %>% filter (q_value < 0.05) %>%
  select(gene_short_name, term, q_value, estimate)
```



## Graph auto-correlation analysis, to compare clusters

WARNING: below chunk will take around 20 mins to run (In my laptop atleast). IF we have the RDS file we can skip below chunk
```{r, message=FALSE, echo=FALSE}
#deg_graph_test_res <- graph_test(cds, neighbor_graph="knn") # Takes long time to run, RDS is stored
```


```{r, message=FALSE, echo=FALSE, fig.height=10}
deg_graph_test_res <- readRDS(file = "deg_graph_test_res.rds")

pr_deg_ids <- row.names(subset(deg_graph_test_res, q_value < 0.05))

gene_module_df <- find_gene_modules(cds[pr_deg_ids,], resolution=1e-2)


cell_group_df <- tibble::tibble(cell=row.names(colData(cds)), 
                                cell_group=clusters(cds))         # Can either use function clusters or partitions
agg_mat <- aggregate_gene_expression(cds, gene_module_df, cell_group_df)

row.names(agg_mat) <- stringr::str_c("Module ", row.names(agg_mat))
colnames(agg_mat) <- stringr::str_c("clusters ", colnames(agg_mat))

pheatmap::pheatmap(agg_mat, cluster_rows=TRUE, cluster_cols=TRUE,
                   scale="column", clustering_method="ward.D2",
                   fontsize=6)


plot_cells(cds, 
           genes=gene_module_df %>% filter(module %in% c(18,28,22)),
           group_cells_by="cluster",
           color_cells_by="cluster",
           show_trajectory_graph=FALSE, cell_size = 2)

gene_module_df[gene_module_df$module == 18,]
```

## Trajectory graph and Pseudotime

```{r, message=FALSE, echo=FALSE}
cds <- learn_graph(cds)
cds <- order_cells(cds)     #Choose " root" node
plot_cells(cds, color_cells_by = "pseudotime",
            cell_size = 1)


```

WARNING: below code takes 15-20 mins to run (In my laptop atleast)
```{r}
#pseudotime_cds_pr_test_res <- graph_test(cds, neighbor_graph="principal_graph")
```

```{r, message=FALSE, echo=FALSE, fig.height=10}
pseudotime_cds_pr_test_res = readRDS(file = "pseudotime_cds_pr_test_res.rds")

pr_deg_ids <- row.names(subset(pseudotime_cds_pr_test_res, q_value < 1e-20)) #Absurdly low q-values??!

plot_cells(cds, genes = "Vta1", cell_size = 1)

gene_module_df <- find_gene_modules(cds[pr_deg_ids,], resolution=0.001) #Not working at certain resolutions, get error in leidenbase or contrasts..

cell_group_df <- tibble::tibble(cell=row.names(colData(cds)), 
                                cell_group=clusters(cds))
agg_mat <- aggregate_gene_expression(cds, gene_module_df, cell_group_df)
row.names(agg_mat) <- stringr::str_c("Module ", row.names(agg_mat))
pheatmap::pheatmap(agg_mat,
                   scale="column", clustering_method="ward.D2")

plot_cells(cds, 
           genes=gene_module_df %>% filter(module %in% c(2)),
           group_cells_by="cluster",
           color_cells_by="cluster",
           show_trajectory_graph=FALSE, cell_size = 1) # Since trajectory graph goes from cluster to cluster, how can we be sure this isnt same as diffgene exp for cluster/

gene_module_df[gene_module_df$module == 2,]
```
Now we are looking at each branch to analyze

```{r}
 # analyzing branches for pseudotime analysis

cds_Serum_branch <- choose_cells(cds)
cds_rest_branch <- choose_cells(cds)


#cds_serum_pr_test_res <- graph_test(cds_Serum_branch, neighbor_graph="principal_graph")
#cds_rest_pr_test_res <- graph_test(cds_rest_branch, neighbor_graph="principal_graph")

cds_serum_pr_test_res <- readRDS("cds_serum_pr_test_res.rds")
cds_rest_pr_test_res <- readRDS("cds_rest_pr_test_res.rds")

pr_deg_serum_subset_ids <- row.names(subset(cds_serum_pr_test_res, q_value < 0.05))
pr_deg_rest_subset_ids <- row.names(subset(cds_rest_pr_test_res, q_value < 0.05))


gene_module_serum_df <- find_gene_modules(cds_Serum_branch[pr_deg_serum_subset_ids,], resolution=0.001)
gene_module_rest_df <- find_gene_modules(cds_rest_branch[pr_deg_rest_subset_ids,], resolution=0.001)


# Organize modules by similarity (using hclust) over the trajectory
## Serum
agg_mat_serum <- aggregate_gene_expression(cds_Serum_branch, gene_module_serum_df)
module_dendro_serum <- hclust(dist(agg_mat))
gene_module_serum_df$module <- factor(gene_module_serum_df$module, 
                                levels = row.names(agg_mat_serum)[module_dendro_serum$order])
## Rest
agg_mat_rest <- aggregate_gene_expression(cds_Serum_rest, gene_module_rest_df)
module_dendro_rest <- hclust(dist(agg_mat))
gene_module_rest_df$module <- factor(gene_module_rest_df$module, 
                                      levels = row.names(agg_mat_rest)[module_dendro_rest$order])


## removing NAs
gene_module_serum_df = gene_module_serum_df[complete.cases(gene_module_serum_df), ]
gene_module_rest_df = gene_module_rest_df[complete.cases(gene_module_rest_df), ]


```

Now we are plotting them in different ways

```{r}

### plotting serum heatmap
cell_group_df_serum <- tibble::tibble(cell=row.names(colData(cds_Serum_branch)), 
                                cell_group=clusters(cds_Serum_branch))
agg_mat_serum_heatmap <- aggregate_gene_expression(cds_Serum_branch, gene_module_serum_df, cell_group_df_serum)
row.names(agg_mat_serum_heatmap) <- stringr::str_c("Module ", row.names(agg_mat_serum_heatmap))
pheatmap::pheatmap(agg_mat_serum_heatmap,
                   scale="column", clustering_method="ward.D2")



# Plotting modules
plot_cells(cds_Serum_branch,
           genes=gene_module_serum_df,
           label_cell_groups=FALSE,
           show_trajectory_graph=FALSE,
           cell_size = 2) #  What to infer from this plot, how to draw conclusions? genes seem to show trajectory or polarization within cluster, not between them



gene_module_serum_df[gene_module_serum_df$module == 7, ]


```



From the modules we can select some interesting genes

```{r}
interest_genes <- c("C1qb", "Apoe")
interest_cds <- cds[rowData(cds)$gene_short_name %in% interest_genes,]

plot_genes_in_pseudotime(interest_cds, cell_size = 1, color_cells_by = "pseudotime",
                         min_expr=0.5)
plot_cells(cds, genes = interest_genes, cell_size = 1 )

```




